SOM: Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe

### 2019-09-11 analysis/paper/SOM.Rmd
formatting before submission

---
title: "Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe"
author:
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::word_document2:
    fig_caption: yes
    reference_docx: "../templates/template2.docx" 
bibliography: LGMTechnology.bib
csl: "../templates/plos-one.csl"
abstract: |
  Climate changes that occurred during the Last Glacial Maximum (LGM) had significant consequences in human eco-dynamics across Europe. Among the most striking impacts are the demographic contraction of modern humans into southern refugia and the potential formation of a population bottleneck. In Iberia and southern France transformations also included the occurrence of significant technological changes, mostly marked by the emergence of a diverse set of bifacially-shaped stone projectiles. The rapid dissemination of bifacial technologies and the geographical circumscription of specific projectile morphologies within these regions have been regarded as evidence for: (1) the existence of a system of long-distance exchange and social alliance networks; (2) the organization of human groups into cultural facies with well-defined stylistic territorial boundaries. However, the degree and modes in which cultural transmission have occurred within these territories, and how it may have influenced other domains of the adaptive systems, remains largely unknown. Using southern Iberia as a case-study, this paper presents the first quantitative approach to the organization of lithic technology and its relationship to hunter-gatherers territorial organization during the LGM. Similarities and dissimilarities in the presence of metric and morphological data describing lithic technologies are used as a proxy to explore modes and degrees of cultural transmission. Statistical results show that similarities in technological options are dependent on the chronology and geographical distance between sites, and confirm previous arguments for the organization of LGM settlement in Southern Iberia into discrete eco-cultural facies.

keywords: |
  Last Glacial Maximum; Solutrean; Lithic Technology; Iberian Peninsula
---

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  dpi = 300,
  fig.width = 9,
  fig.height = 6,
  fig.path = "../figures/"
  )

library(knitr)
library(rmarkdown)
library(git2r)
library(ggplot2)
library(ggpubr)
library(readr)
library(plyr)
library(dplyr)
library(tidyr)
library(tibble)
library(tab)
library(forcats)
library(FactoMineR)
library(factoextra)
library(multcompView)
library(RcmdrMisc)
library(gridExtra)
library(DescTools)
library(WRS2)
library(reshape2)

@@

```{r ANOVA_comparisons, echo = FALSE}

# List of comparisons to use with grouping labels (dash with no spaces must be the separator)
comparisons <- c("AMB II-AMB IV", "AMB II-AMB VI", "AMB II-PAP 4'00",
                 "AMB II-PAP 4'75", "AMB II-PAP 5'25", "AMB II-VALM",
                 "AMB II-VB A", "AMB II-VB B", "AMB II-VB C",
                 "AMB IV-AMB VI", "AMB IV-PAP 4'00", "AMB IV-PAP 4'75",
                 "AMB IV-PAP 5'25", "AMB IV-VALM", "AMB IV-VB A", "AMB IV-VB B",
                 "AMB IV-VB C", "AMB VI-PAP 4'00", "AMB VI-PAP 4'75",
                 "AMB VI-PAP 5'25", "AMB VI-VALM", "AMB VI-VB A",
                 "AMB VI-VB B", "AMB VI-VB C", "PAP 4'00-PAP 4'75",
                 "PAP 4'00-PAP 5'25", "PAP 4'00-VALM", "PAP 4'00-VB A",
                 "PAP 4'00-VB B", "PAP 4'00-VB C", "PAP 4'75-PAP 5'25",
                 "PAP 4'75-VALM", "PAP 4'75-VB A", "PAP 4'75-VB B",
                 "PAP 4'75-VB C", "PAP 5'25-VALM", "PAP 5'25-VB A",
                 "PAP 5'25-VB B", "PAP 5'25-VB C", "VALM-VB A", "VALM-VB B",
                 "VALM-VB C", "VB A-VB B", "VB A-VB C", "VB B-VB C")

@@

# Introduction

Centered around c. 21 ka cal BP, the Last Glacial Maximum (LGM) was a global climate event characterized by the maximum expansion of the ice sheet over Scandinavia and northern Europe, along with generally cold and arid conditions in central and western Europe [@clark_last_2009; @hughes_timing_2013; @mix_environmental_2001].

Climatic changes occurred during this period had substantial repercussions in the geography, technology, and social behavior of human populations in Europe. Among the most drastic impacts are: (1) the contraction of the human range to southern regions of Europe that served as refugia, such as the Iberian and the Italian peninsulas [@banks_human_2008; @bocquet-appel_population_2000; @bradtmoller_repeated_2012; @burke_exploring_2014; @derrico_neandertal_2003; @jochim_late_1987; @straus_riera_1986; @straus_southwestern_1991]; (2) the consequent genesis of a bottleneck scenario in human genetic diversity [@barbujani_evidence_1998; @keinan_measurement_2007; @torroni_signal_2001]; (3) and, in southwestern Europe, the advent of a set of new technologies for the manufacture of lithic projectiles and knives using bifacial retouch.

This latter feature is one of the main elements traditionally used to define the so-called Solutrean techno-complex [@smith_solutreen_1966]. The Solutrean is a phenomenon geographically limited to southwestern France and the Iberian Peninsula, presenting a moderately short chronological range (c. 25–19 ka cal BP) when compared to other pan-European techno-complexes and representing, to a certain extent, an outlier within the common background persisting throughout the Upper Paleolithic in western Europe.

For its uniqueness and rough coincidence with the LGM, the Solutrean record often emerges as a promising context for examining the mechanisms of formal and chronological change of lithic tools during the late Pleistocene. Further, it has been recognized as an excellent opportunity to test the possible links between the development of specific technical innovations and social behavior under a particular climate setting [@derrico_limpact_2006; @renard_continuity_2011; @villaverde_upper_1998].

In fact, besides the long-standing recognition of the several types of bifacial implements as secure chrono-cultural markers, their presence across southwestern Europe, and the notion that they may convey complex symbolic communication, has inspired many theories and models on social networks, territorial organization, demographic changes, and the subsistence of hunter-gatherers.

Two perspectives have been particularly relevant over time. One that denotes the existence of a system of long-distance exchange and alliance networks, based on the identification of uniform techno-economic patterns across all of southern France and Iberia (including the use of bifacial technologies, but also the presence of similar patterns in raw material selection, systematic use of heat treatment and stylistic similarities in art) [@aubry_solutrean_2003; @straus_quarter-century_2000; @zilhao_seeing_2013]. The other, in contrast, based on the recognition of spatially-patterned bifacial typological traditions (e.g. Cantabrian vs Mediterranean) [@smith_solutreen_1966; @straus_pointes_1977; @tiffagom_pierre_2006; @zilhao_o_1997], and justified by the existence of environmental/ecological constraints [@banks_investigating_2009; @burke_exploring_2014], seem to have resulted in the delineation of stylistic territorial boundaries and an inherent recognition of social group ownership by populations inhabiting rather distant regions [@bicho_fashion_2009; @cascalheira_influencia_2013].

Overall, these patterns attest a complex form of sharing behaviors occurring across different scales during the LGM. From a demographic point of view, this implies the existence of a very dynamic population circulating over significantly large territories, resulting in the permutation of different kinds of information, including, very likely, genetic exchange [@straus_quarter-century_2000; @straus_solutrense_2013].

Since most models on LGM territoriality, however, are based on the study of bifacial and shouldered projectiles, there is a significant lack of information on how cultural transmission, on the one hand, and regional ecology, on the other, may have influenced other domains of the adaptive systems. Notably, it is still largely unknown to what extent technological organization and, mostly, mental templates for lithic reduction sequences, were shaped by territorial organization during this period. Are similarities within regions only expressed at the symbolic/stylistic level, with communities sharing projectile morphologies and artistic elements as a way to express belonging to a common territory? Or, on the other hand, did the contexts in which long-distance contacts seem to be present also exhibit similarities at other levels of the technological system, including, for example, central tendencies in the lithic reduction sequences?

Answering these questions becomes particularly relevant given that bifacial tools, albeit important, are just a singular component in LGM technological structures. Solutrean industries are, in reality, mostly composed of a non-bifacial techno-typological substrate [@sinclair_technique_1995; @straus_original_1990], comparable to that present at any other Upper Paleolithic techno-complex [@bar-yosef_upper_2002], that formed the basis for the production of a wide range of blanks and tools.

In this context, this paper provides the first quantitative comparison between the non-bifacial components of LGM lithic assemblages from a set of sites located at geographically distant regions but within the same stylistic/cultural territory. Using southern Iberia as a case-study, specific technological domains of the assemblages are statistically compared to explore scenarios of territoriality and the organization of technology across the LGM.

## LGM archaeology in southern Iberia

In Iberia, four cultural facies are distinguished based on the distribution of bifacial and/or shouldered point types during the LGM (Fig. XXXX). The southern part of the Peninsula encompasses two of them, known as Mediterranean and Portuguese facies  [@tiffagom_pierre_2006]. Both have in common the presence of abruptly retouched shouldered points [@villaverde_piezas_1981-1] and stemmed and winged projectiles [@fullola_les_1985; @pericot_cueva_1942]. They are, however, separated by the fact that at some sites from central Portugal Cantabrian-type shouldered points (with bifacial and flat retouch) are also documented [@zilhao_o_1997]. This separation between central Portugal and the remaining territory also became evident in previous analysis of stemmed and winged projectiles morpho-metrics. Points from central Portugal are generally larger, in both length and width  [@bicho_fashion_2009; @cascalheira_tecnologia_2010], and present significantly different stem shapes [@schmidt_solutrean_2015] than the ones from southern Portugal and Mediterranean Spain. While these patterns seem to indicate some distinction between regions, the exclusive distribution of both projectile morphologies below 40º N latitude conventionally attests the validity of using southern Iberia as a broad cultural/stylistic unit. Artistic particularities, mostly from mobile art, also corroborate this territorial homogeneity [@bicho_solutrean_2012; @bicho_upper_2007; @simon_vallejo_primeras_2012; @villaverde_arte_1994; @zilhao_plaquette_1988].

```{r lgmmap, echo=FALSE}

@@

In chronological terms, according to the traditional models [e.g.,  @jorda_solutrense_1955; @munoz_ibanez_puntas_2000; @tiffagom_pierre_2006; @zilhao_o_1997; @zilhao_seeing_2013; @zilhao_sequence_1994], backed shouldered projectiles and stemmed and winged points are only characteristic of the later phases of the Solutrean techno-complex (i.e., Upper Solutrean and Solutreo-Gravettian). However, this interpretation has been questioned in a recent Bayesian analysis of the currently available radiocarbon data [@cascalheira_chronological_2015], calling into doubt the status of those type-fossils as precise temporal markers, and confirming their presence throughout most of the LGM time-span (i.e., c. 25-19 ka cal BP).

LGM settlement of southern Iberia is marked by the existence of more or less dense clusters of sites, that are relatively isolated from one another by geographical hiatus where the complete absence or the presence of very few sites might testify human passage but not likely intensive occupation [@aura_cueva_2009; @cascalheira_vale_2012]. Caves and rock-shelters are better represented than open-air locations, as are multi-component sites over single horizon ones [@cascalheira_chronological_2015; @cascalheira_influencia_2013]. There is evidence for the existence of different types of sites, including long-term and seasonal residential camps, shorter specialized occupations, lithic cache sites, and possible aggregation sites [sensu @conkey_identification_1980]. The number of occupied sites is significantly higher than during the early Upper Paleolithic and, although they tend to be located in coastal or near-coastal settings, inland camps are also represented, mainly as single-occupation localities [@schmidt_rapid_2012; @straus_upper_2000].

Rather problematic is the quality of the available data from most of these sites. More than half of the occurrences are surface or mixed contexts, suffering from a number of limitations (e.g. lack of organic materials, presence of small and/or highly truncated lithic assemblages, and occurrence of thick palimpsest deposits - @aubry_analyse_2013; @cascalheira_chronological_2015; @cascalheira_influencia_2013; @cascalheira_o_2013; @zilhao_seeing_2013].

Probably as a result of this scenario, technological analyses of LGM lithic materials are (as well as faunal analysis) very limited in southern Iberia. These are mostly restricted to intra-site descriptions and evaluation of the assemblages [@aura_cueva_2009; @cascalheira_tecnologia_2010; @cascalheira_vale_2012; @ripoll_lopez_cueva_1988; @tiffagom_pierre_2006] and, very rarely, presenting inter-site or regional comparisons using technological patterns for the construction of clear paleoanthropological testable hypothesis [see @barton_glacial_2013 as an exception].

Two rare examples are the studies conducted by Zilhão [-@zilhao_o_1997] and Schmidt [-@schmidt_solutrean_2015]. The first, focusing on central Portugal, uses assemblage composition and raw material circulation to report a scenario of logistically organized resource-procurement strategies within the region. Body adornments distribution and technological data allowed the author to further suggest the organization of the region into smaller territories corresponding to different ethnic groups, sharing specific traditions but all inserted within the same ethnic boundary.

The second study addresses Iberia as a whole and compares the technological organization behind the manufacture and use of the different types of lithic projectiles. Previously referred significant differences between northern and southern Iberia are corroborated in this study, and the main outcome is that synchronic variability of Solutrean points between those regions does not only relate to stylistic variations, but to regional differences deeply rooted in the distinct technological strategies of tool making and tool use. Corroborating Zilhão [-@zilhao_o_1997] the author states that: “the technological design of southern Iberian points (…) encompasses criteria of toolkits ascribed to logistical mobility of collectors, with a high functional specialization of the points, and evidence at `base camps’ for batch production” [@schmidt_solutrean_2015].

Both models seem to confirm the scarce data coming from faunal studies [@manne_intensive_2012; @yravedra_zooarqueologiy_2007] that demonstrate a seasonal movement of residential sites between coastal areas and the mountainous hinterland, and possibly a perennial occupation of few major sites across southern Iberia [see also @barton_glacial_2013].

In sum, paleoanthropological models for LGM hunter-gatherer adaptations in southern Iberia are currently rather limited [contra northern Iberia – see e.g., @straus_new_1976; @straus_quarter-century_2000; @straus_study_1975] and mostly based on the typo-technological analysis of bifacial and shouldered projectiles or other “style-embedded” elements. Some of these elements report, on the one hand, significantly homogeneous patterns across the region and, on the other, regional particularities that suggest the organization of settlement into smaller socio-cultural territories.

# Theoretical framework for cultural transmission processes

The diffusion of specific concepts across southern Iberia, as rapidly as it seems to have occurred during the LGM, may have resulted from an array of very diverse contact behaviors. Ethnoarchaeological data [@wiessner_reconsidering_1984; @wiessner_risk_1982; @wiessner_style_1983] has demonstrated that the process of social interaction between hunter-gatherer communities can vary between episodes in which people have little contact to each other’s domestic space (with gatherings occurring at the edge of their respective ranges), to events in which marriage partners are exchanged, and previously foreign individuals become fixtures in each other’s residential lives [@tostevin_seeing_2013; @tostevin_social_2007]. Contacts occurring anywhere between these two extremes of the contact scale result in dissimilar access to information, and to different types of information.

In the case of LGM projectiles and remaining shared concepts, both hypotheses are viable to have occurred. The design concepts of shouldered and stemmed and winged projectile points may have been disseminated in intimate cultural environments with access to the full production sequences, or it may, on the contrary, be the outcome of independent innovation by exposure to curated tools and not necessarily to the complete technological mechanisms. Within this latter scenario, although it is possible that an individual could independently develop some of the behaviors of point production from examining a finished tool, it is rather unlikely that he/she would use the same procedures, techniques, and methods, during the knapping process. Thus, understanding the complete production process is of great importance to distinguish between modes but mostly degrees of social contact, and how these vary across time. Focusing exclusively on information coming from projectiles morphologies, as well as on any other class of traditionally considered style-embedded items (i.e., portable art, ornaments, lithic retouched tools), will only allow to explore the patterns of presence/absence of contacts and not the possible nature of those contacts. Further, using LGM bifacial reduction sequences and projectile manufacture in southern Iberia to shed light on the technological processes is a somehow hampered task. Patterns of segregation of the production sequence, which comprise shaping of preforms at sites close to rich raw material sources on the one hand, and export of preforms and finished points to other locations on the other hand [@schmidt_solutrean_2015; @tiffagom_pierre_2006], can severely truncate the variability of technological procedures. Since, however, the available information on point production attests, in a large number of cases, that bifacial points were obtained from blanks produced within non-bifacial reduction sequences [@cascalheira_tecnologia_2010; @tiffagom_solutrense_2005; @zilhao_o_1997], a reasonable approach is to look at the non-bifacial technology of the assemblages and check for patterns of similarity/dissimilarity in the distribution of specific methods and concepts within the reduction sequence continuum.

Such kind of analysis requires, however, a theoretical and methodological framework that would allow the integration of both social interactions and ecological background as shapers of stone knapping patterns across space and time. While there are many ethnoarchaeological and theory-based studies addressing the meaning of morphological variations of lithic tools as indicators of style and territorial differentiation, very few attempts have been made to use technological processes as proxies for cultural transmission processes. The middle-range theory presented by Tostevin [-@tostevin_seeing_2013; -@tostevin_social_2007], for studying the material evidence of social contact between Neanderthals and Anatomically Modern Humans using lithic technology, is perhaps one of the most compelling and sophisticated approaches. Based on a body of literature on artefact style [@jochim_building_1995; @jochim_unified_1995; @lechtman_style_1975; @sackett_style_1990; @wiessner_reconsidering_1984; @wiessner_seeking_1997], optimization studies [@bird_behavioral_2006; @boyd_culture_1985; @broughton_evolutionary_1999; @kuhn_evolutionary_2004] and experimental knapping [@dibble_new_1981; @pelcin_effect_1997; @pelcin_effect_1997-1; @pelcin_formation_1997], Tostevin’s model seeks to identify whether variation in lithic material culture style reflects one or more social processes, once other sources of variability have been dismissed.

The present paper adopts the same body of theory to inform a technology-based approach (see methods below).

Ethnography has shown that technological gestures are not just individual action sequences with clearly defined goals, but suites of distributed behaviors reflecting complex systems that have been learned and assimilated as bodily techniques. In this context, flintknapping processes are considered “technological performances which can only be effectively taught and learned by observation of the body movements, the handling of core and percussor, and the flake by flake strategizing on how to exploit core volumes” [@tostevin_social_2007, pp. 344]. Lithic reduction is a directional and irreversible process characterized by a sequence of mechanical relationships. While these relationships are influenced by multiple contextual, aspects, such as the quality of the available raw material, identifying different exploitation patterns within these mechanical relationships are thought to reflect a purposeful choice of particular modes of technological manipulation present in an assemblage out of an equifinal pool of alternative possibilities - what Sackett [-@sackett_approaches_1982; -@sackett_style_1986; -@sackett_style_1990] have called isochrestic variation .

Based on this theoretical approach, it is assumed here that the characterization and comparison of the central tendencies of lithic attributes distribution within specific and comparable clusters of technological decisions can be used as indicators of behavior differentiation. In this way it is possible to assess how knappers responsible for different assemblages solve problems, react to different circumstances and produce artifacts according to specific learned methods, resulting in the variability detected among assemblages. It is also assumed, thus, that technological and technical choices, whether conscious or not, taken by knappers are integral components of cultural life and are “largely dictated by the technological traditions within which they have been enculturated as members of the social groups that delineate their ethnicity”  [@sackett_style_1990, pp. 33]. The same choices can though be subject to revision as a result of changes in the patterns of social interaction between groups.

Consequently, the greater the similarities recorded between assemblages, the greater the probability of concepts exchange to have occurred between knappers responsible for different archaeological assemblages, most likely in a context of social intimacy, with full access to technological conceptual schemes. On the contrary, if dissimilarities are more evident, than either technological practices transmission occurred in the context of a low degree of social intimacy, and/or ecological and cultural [local social learning - @premo_spatial_2011; @scholnick_spatial_2012] regional constraints did not allow the generalization of basic technological patterns.

This study builds, therefore, upon the theoretical assumption that the amount of lithic technological knowledge, learned and adopted by hunter-gatherers, depends upon how and where learning occurs and may reflect differential access to particular enculturating environments, and thus "are determined by the mechanics of how the traits are transmitted from individual to individual and by the stochastic events associated with these" [@shennan_isolation-by-distance_2015, pp. 103].

The model expectations for different modes of cultural transmission during the LGM in southern Iberia based on technological analysis of lithic attributes presented in this paper are summarized in Table 1. Similar approaches using lithic technology as a proxy for demography and cultural transmission have been applied in several other studies [e.g. @henrich_demography_2004; @scerri_unexpected_2014; @shennan_demography_2001; @tostevin_seeing_2013].

Some caveats are associated with the expectations presented since they result from a complex interplay among patterns of inheritance, interaction, and local adaptation. Whereas equifinality and the restricted options inherent to leptolithic technologies may result in similar technological trends, even within communities with none or very little contact; aspects such as chronology, functional variability, and ecological constraints (e.g., raw material variability) may influence the presence of significantly different lithic attributes among populations with a high degree of interaction. Most of these alternative explanations are related to several adaptive behaviors as they are organized across different temporal and geographical scales, within which technological organization is composed of a set of problem-solving processes that are responsive to conditions created by the interplay between hunter-gatherers and their environment [@nelson_study_1991]. Nonetheless, even in the presence of rather different assemblages due to specificities related with contextual aspects there is, based on the theoretical framework presented above, and based on the assumption that all behavior may, in one sense or another, be regarded as purposeful, always a bound between technological practices and the definition of a particular social/cultural unit. Contextual constraints were, however, not downplayed in the course of the analysis and interpretation of results obtained in this study. Additionally, the focus on a small-scale analysis, taking full advantage of statistical methods, and targeting meaningful comparability between contexts were purposefully used in this approach to reduce the negative impact of possible errors and uncertainty.

# Sites and materials

Sites included in this study were selected to ensure, as much as possible, chronological correspondence, spatial variability, and the integrity and comparability of contexts covering the whole region under study. The set of selected sites comprises one open-air context from central Portugal (Vale Almoinha), one rock shelter from southern Portugal (Vale Boi), and two sites - one rock shelter and one cave (respectively, Ambrosio and Parpalló) - from Mediterranean Spain Figure \@ref(fig:lgmmap).

Vale Almoinha (VALM) is a single LGM occupation open-air site located in the Portuguese Estremadura. It was first excavated by Manuel Heleno in the late 1940s and relocated and dated by Zilhão in 1986 [@zilhao_estratigrafia_1987]. The site is located in a small inland valley at c. 1.5 km from the present-day coastline. Two radiocarbon dates on charcoal samples are available, dating the archaeological horizon to c. 24-24.5 ka cal BP [@zilhao_o_1997]. Previous analysis of the lithic materials [@zilhao_o_1984; @zilhao_o_1997] revealed an assemblage dominated by chert, some quartzite and residual presence of quartz. Core maintenance products are well represented. Lithic analysis revealed the presence of two distinct approaches, one for the specific production of laurel-leafs and other for the production of flakes and elongated blanks. Laurel-leaf points are the well-most represented tool type, followed by unifacial flat-retouched points, willow-leafs and two, questionable [see @schmidt_solutrean_2015], shouldered points. The overall characteristics of the assemblage and site context allowed Zilhão [-@zilhao_o_1997] to characterize it as a semi-permanent base camp, in which are represented the various phases of the *chaîne operatoire*.

Vale Boi (VB) is a multilayered and multicomponent (open-air and rock shelter) site located in the southwesternmost extreme of Iberia (southern Portugal). The site is situated in the eastern side of a wide valley, 2.5 km inland from the present-day coastline, occupying a wide portion of a relatively steep slope, marked by the presence of a 10-15 m high limestone cliff on the top. LGM occupations are attested across all excavated areas, but the most complete sequence was identified below the boulders resulting from the collapse of a relatively large rock shelter [@bicho_ledge_2012]. Six radiocarbon dates indicate that Solutrean occupations at VB rock shelter started somewhere around 25 ka cal BP and ended shortly before c. 20.5 ka cal BP [@cascalheira_chronological_2015; @cascalheira_vale_2012]. Previous analysis of lithic assemblages revealed the presence of backed shouldered and stemmed and winged points across the sequence of three lithostratigraphic units defined during excavation. Chert, quartz, and greywacke are the most common lithic materials recovered, all provisioned locally, from deposits situated no more than 15-20 km from the site. The density and diversity of findings, together with data from faunal analysis [@manne_early_2014; @manne_intensive_2012], allowed the classification of VB as a seasonal residential camp.

Ambrosio (AMB) rock shelter is located in the eastern part of Andaluzia, positioned at an elevation of 1060 meters above sea level, and distant 60 km from the Mediterranean coast. The shelter has a south-southwest orientation and a triangular shape whose entry measures c. 39 meters wide by c. 19 meters in height, and extends c. 17 meters into the Miocene limestone [@ripoll_lopez_cueva_1988]. The site was excavated by Eduardo Ripoll Perelló between 1958 and 1964 and by Sergio Ripoll Lopez from 1982 onwards, and it is mostly known for the Solutrean parietal art [@ripoll_lopez_arte_1994]. The Pleistocene sequence is marked by a series of Solutrean occupations organized by the excavators into three distinct horizons, separated by clast-supported levels with fewer materials. Recent AMS radiometric results contradict previous standard dates, indicating that the most recent occupation occurred at c. 24 ka cal BP. Backed shouldered and stemmed and winged points are present throughout levels II and IV but completely absent from level VI. As in the VB case, this site has been classified as a seasonal residential camp.

The Parpalló (PAP) cave is located in a wide valley, some 6 km from the present coastline. At c. 600 meters above sea level the cavity entrance is marked by a relatively narrow breach, which gives access to the main room with c. 80 m2. The first work was by Breuil in 1913. Later, between 1920 and 1930, L. Pericot undertakes a series of campaigns resulting in the almost complete excavation of the archaeological deposits [@pericot_cueva_1942]. With close to 8.5 meters of thickness, the stratigraphy at PAP comprises Gravettian, Solutrean and Magdalenian occupations. Unfortunately, excavation was done by subdividing the deposits into artificial spits of 20-25 centimeters and screening all sediments using fairly large meshes [@tiffagom_pierre_2006]. Standard radiocarbon ages are available for levels 5’00-4’75 and 4’25-4’00, that, after Bayesian analysis date the horizons to c. 22.5-23 ka cal BP [@cascalheira_chronological_2015]. Shouldered and stemmed and winged points are present throughout the sequence but clearly are better represented in levels between 5.25 and 4.25 meters in depth [@munoz_ibanez_puntas_2000; @tiffagom_pierre_2006]. This site was possibly used as a residential space, although the remarkable and unique amount of recovered engraved slabs [@villaverde_arte_1994] might indicate the presence of some symbolic or ritualistic functionality.

Given the size of the Spanish assemblages, materials were sampled in the case of AMB from the 1986 campaign [@ripoll_lopez_cueva_1988], and in the case of PAP from sector L [@tiffagom_pierre_2006]. Although randomly selected, the representativeness of these samples within each site is thought to be good, given the large portion of the excavation area they represent.

For the three karstic contexts, a diachronic subdivision of the assemblages was also implemented based on the available lithostratigraphic information and previous analysis of the materials [@cascalheira_vale_2012; @munoz_ibanez_puntas_2000; @ripoll_lopez_cueva_1988; @tiffagom_pierre_2006].  This subdivision resulted in a total of 10 contexts, representing, as much as possible, the chronological and geographical variability of LGM adaptations in southern Iberia.

All sites are radiocarbon dated, yet direct comparisons between contexts are problematic given the sometimes very large standard deviations of the results.

Further, all four chosen sites are located near raw material sources, and thus all assemblages have been characterized as primary flaking assemblages, containing features such as a reasonable number and diversity of core maintenance pieces and large cortical flakes. A significant difference between sites is in the diversity of raw materials used. Although chert is always dominant, in Portuguese assemblages a greater diversity of materials is present. In the particular case of VB, non-siliceous raw materials were used not as alternatives to chert but for very specific functional purposes [@cascalheira_vale_2012]. This pattern is consistent with that recorded for nearly all upper Paleolithic sites in the western Atlantic facade of Iberia even when good-quality chert is available within walking distance [@pereira_variation_2013]. To avoid a possible impact on assemblage comparability caused by the use of different raw materials, only chert artifacts were selected for this study. This restriction was also imposed by the composition of the assemblages since for both Spanish sites it was not possible to identify more than a handful of artifacts made on other materials than chert, a fact that is also evident in previous publications for Ambrosio [@ripoll_lopez_cueva_1988].

# Methods

Complete cores and blanks were analyzed using a set of technological variables and attributes, specifically chosen to allow the isolation of separate sources of variability, including raw material constraints and learned behaviors (see Supplementary Online Material [SOM] Tables S1 and S2 for a complete list of attributes and measurement techniques).

Following the theoretical background of Tostevin [-@tostevin_seeing_2013; -@tostevin_social_2007], Scerri [-@scerri_aterian_2013] and Scerri et al. [-@scerri_unexpected_2014], in this study variables were organized into comparable heuristic domains (i.e. Platform Maintenance, Core Exploitation, Dorsal Convexities Management, Elongated blanks Production, Core Morphology, and Core Use) of the technological sequences, to achieve a comprehensive structure of similarity and dissimilarity between assemblages. The main goal of using these domains was to isolate different sources of variability by focusing on specific sets of interrelated variables that correspond to interdependent knapping actions. Breaking down the reduction sequences into comparable clusters makes possible to account for possible biased samples and “broken” sequences, not assuming that the knapping process was restricted to a specific timeframe, or the work of a single individual [@scerri_unexpected_2014].

The chosen variables, and respective combinations, were based on the traditional parameters used in lithics attribute analysis [see e.g., @andrefsky_jr_lithics_2005; @inizan_technology_1999] and the results of experimental studies available in the literature [see e.g., @cotterell_formation_1987; @dibble_effect_1995; @dibble_introducing_2009; @dibble_new_1981; @pelcin_effect_1997; @pelcin_effect_1997-1; @pelcin_formation_1997; @pelcin_threshold_1998; @rezek_relative_2011]. For example, controlled knapping experiments show that blank dimensions and shape are significantly affected by variations in exterior platform angles, platform depths and platform preparation (e.g., faceting) [@dibble_introducing_2009; @rezek_two_2018-2]. Thus, by considering variables such as platform type and blank elongation (length/width ratio) as a set of interrelated variables, the way knappers differently exploited these variables can be assessed and compared across assemblages. The same is true for other domains, such as the exploitation of dorsal surface convexities or the management of core morphology, all with direct impact on blank morphometrics.

A multivariate statistical approach was used to explore the interaction of variables within each domain. Since most of the attributes available were categorical, a Multiple Correspondence Analysis (MCA) was used to explore the orthogonal dimensions of variability in the data. MCA is well suited to lithic analysis because it allows the isolation of the different sources of variability contributing to the character of each assemblage under study. When used, metric variables were pooled into categories using a K-means clustering algorithm. To reduce the amount of ‘noise’ in the overall variability, all categorical attributes whose representation was less than 5% within each variable were also lumped together into a category named ‘Other’, before any of the analysis were performed.

Following Scerri et al. [-@scerri_unexpected_2014] the scores and loadings of the most relevant dimensions calculated by the MCA were used to identify and explore the relative contributions of aspects such as reduction intensity, raw material constraints, or particular knapping styles. To analyze the main sources of variability behind the technological character of each assemblage separately, and to assess the similarities and differences between contexts, the individual scores of each relevant component of the MCA were statistically tested by applying an analysis of variance (ANOVA). ANOVA tests and corresponding post hoc comparisons were run using a percentile t bootstrapped approach for trimmed means (trim level was set at 20%) as implemented by Mair and Wilcox [-@mair_robust_2016]. This approach intended to prevent problems related to heavy-tailed distributions, unequal sample sizes, and possible differences in skewness distributions of the samples. The location of individuals, grouped by assemblages in the multivariate dimensions, was bootstrapped 5000 times in all tests.

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r_2013]. The complete R code used for all the analysis and visualizations contained in this paper is available at XXXXX. To produce those files the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research were followed. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code to produce all tables and figures. To enable maximum re-use, code is released under the MIT license, data as CC-0, and figures as CC-BY, [for more information see @marwick_computational_2016].

# Results

## General characteristics of the assemblages

The total number of artifacts for each context used in this study is presented in Table \@ref(tab:general). The most numerous assemblages are, respectively, PAP 4’00-4’75 (n = 2743) and VB B (n = 1723), and the less numerous are AMB II (n = 540), and VB A (n = 493). Cores represent between two and eight percent of the assemblages. VALM is the site with the highest percentage of cores (c. 8.52%), and a rather low blank to core ratio (c. 11) when compared to the other contexts.

```{r general, echo = FALSE}

>>>>

---
title: "SOM: Territoriality and the organization of technology during the Last Glacial Maximum in southwestern Europe"
author:
  - João Cascalheira (ICArEHB, University of Algarve, Faro, Portugal) jmcascalheira@ualg.pt
date: '`r Sys.Date()`'
output: 
    bookdown::pdf_document2:
      toc: False
      number_sections: False

header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}

---

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE)

library(tidyverse)
library(tab)
library(RcmdrMisc)
library(kableExtra)

@@

@@

## S1 - Blank attribute frequency

```{r blanks_cross_table, echo=FALSE}

var_list <- c("PlatformType", "CrossSection", "Profile", "DorsalPattern", "EdgeShape", "CortexPerc", "Termination", "Elongation", "Flattening")

blanks_tidy <- as.data.frame(blanks_tidy)

blanks_table <- tabmulti(blanks_tidy, "Context", var_list,
                 p.include = FALSE,
                 n.headings = FALSE,
                 bold.varnames = TRUE,
                 bold.colnames = TRUE,
                 ynames = c(PlatformType = "Platform type", CrossSection = "Cross section", Profile = "Profile", DorsalPattern = "Dorsal pattern", EdgeShape = "Edge shape", CortexPerc = "Cortex %", Termination = "Termination", Elongation = "Elongation", Flattening = "Flattening"))

blanks_table <- blanks_table %>% 
  as_tibble() %>% 
  select(" " = Variable, `AMB II`, `AMB IV`, `AMB VI`, `PAP 4'00-4'75`, `PAP 4'75-5'25`, `PAP 5'25-6'25`, `VALM`, `VB A`, `VB B`, `VB C`, Total = Overall)

knitr::kable(blanks_table, "latex", booktabs = TRUE) %>% 
  kable_styling(latex_options = c("scale_down"))

@@

## S2 - Elongated blanks attribute frequency

```{r elongated_cross_table, echo=FALSE}

var_list <- c("PlatformType", "CrossSection", "Profile", "DorsalPattern", "EdgeShape", "CortexPerc", "Termination", "Elongation", "Flattening")

elongated_tidy <- as.data.frame(elongated_tidy)

elongated_table <- tabmulti(elongated_tidy, "Context", var_list,
                 p.include = FALSE,
                 n.headings = FALSE,
                 bold.varnames = TRUE,
                 bold.colnames = TRUE,
                 ynames = c(PlatformType = "Platform type", CrossSection = "Cross section", Profile = "Profile", DorsalPattern = "Dorsal pattern", EdgeShape = "Edge shape", CortexPerc = "Cortex %", Termination = "Termination", Elongation = "Elongation", Flattening = "Flattening"))

elongated_table <- elongated_table %>% 
  as_tibble() %>% 
  select(" " = Variable, `AMB II`, `AMB IV`, `AMB VI`, `PAP 4'00-4'75`, `PAP 4'75-5'25`, `PAP 5'25-6'25`, `VALM`, `VB A`, `VB B`, `VB C`, Total = Overall)

knitr::kable(elongated_table, "latex", booktabs = TRUE) %>% 
  kable_styling(latex_options = c("scale_down"))
